DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-05 manuscript/manuscript.tex
addressing EJM comments

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4298}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

@@

\abstract{
Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

Building a chronogram is not an easy task.
It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.

Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
DateLife features a versioned, open and fully public chronogram database (McTavish et al., 2015) storing age information in a computer readable format (Vos et al., 2012), an automated and programmatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.

@@

The DateLife algorithm is fully implemented using the R language. The latest stable version of the R package \texttt{datelife} -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

\hypertarget{providing-an-input}{%
\subsection{Providing an input}\label{providing-an-input}}

DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004). The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

DateLife accepts scientific names as input.
These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input. If the flag is inactive, DateLife ignores the inclusive taxon names from the input.

Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. DateLife implements TNRS using OpenTree's taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.

@@

A DateLife search consists of matching processed taxon names to tip labels in a chronogram database. Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
These matching pruned chronograms are now referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
New chronograms can be added to Phylesystem by any user and are immediately publicly available. This permits an immediate update of DateLife's chronogram database to include those new data within a run.

@@

Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Basic summary information available to the user is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing your group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
\end{enumerate}

\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon. We recommend that users provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019). Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007), or combine topologies from source chronogras using a supertree approach.

\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT, Katoh, Asimenos, \& Toh, 2009). Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).

\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reonstruction will fail.

\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}. Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.

\emph{\textbf{Summarizing congruified ages.--}} Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
into a single summary matrix using different methods.
Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics.

Ages per taxon pair can be used as calibrations to date the chosen tree topology.

\emph{\textbf{Dating the tree topology.--}}
By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology. BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data. This minimizes age variance in the resulting chronogram (Campbell O. Webb, Ackerly, \& Kembel, 2008).
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.

@@

Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb et al., 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.

@@

At the time of writing of this manuscript (Apr 14, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
This observation is often encountered in the literature
(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021). For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016). Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version is available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
Sanchez-Reyes, L. L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M. (2022). {datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}. \emph{R Package Version 0.6.2}. Retrieved from \url{https://doi.org/10.5281/zenodo.593938}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). The age of flowering plants is unknown.

>>>>

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4540}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\shorttitle{DateLife: revealing the dated Tree of Life}

@@

\abstract{%
Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

Building a chronogram is not an easy task.
It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.

Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

\hypertarget{processing-an-input}{%
\subsection{Processing an input}\label{processing-an-input}}

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004). The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial specific. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string. If the flag is inactive, DateLife excludes the taxon names above the species level from the input.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.

@@

DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (\url{https://tree.opentreeoflife.org/curator}). This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.

A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
\end{enumerate}

\emph{\textbf{Choosing a topology.--}}
DateLife requires a tree topology to summarize age data upon.
We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

\emph{\textbf{Dating the topology.--}}
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.

For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb, Ackerly, \& Kembel, 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
BLADJ does not incorporate branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes.
\textgreater{} NOTE: do you know how it decides which one to drop if two calibrations conflict? The parent or the child? It would drop both. It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
(Stephen A. Smith \& O'Meara, 2012).
The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
\textgreater{} NOTE: is this true?? Yes

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

At the time of writing of this manuscript (May 06, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
This observation is often encountered in the literature
(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021). For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016). Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.

@@

\texttt{datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{datelife}'s R package stable version is available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J. A., Mozzherin, D., Rees, T., \ldots{} Enquist, B. J. (2013). {The taxonomic name resolution service: An online tool for automated standardization of plant names}. \emph{{BMC Bioinformatics}}, \emph{14}(1). \url{https://doi.org/10.1186/1471-2105-14-16}

@@

\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
Sanchez-Reyes, L. L., \& O'Meara, B. (2022). {\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}. \emph{R Package Release V0.2.2}. Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}

\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
Sanchez-Reyes, L. L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M. (2022). {\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}. \emph{R Package Release V0.6.2}. Retrieved from \url{https://doi.org/10.5281/zenodo.593938}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
